---
title: "Florida Medical Professional Liability "
output:
#  pdf_document:
#    toc: yes
#    toc_depth: '3'
 html_document:
    toc: yes
    toc_depth: '3'
    number_sections: true
---
*This file atarts a fit of the Florida Medical Professional Liability Claims data using **R** code.*

```{r warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE}
library(VGAM)
library(copula)
library(GB2)
library(statmod)
library(numDeriv)
library(Hmisc)
```

# Data


```{r}
SampleData <- read.csv("FloridaMPLExcess2MM.csv", header=T, na.strings=c("."), stringsAsFactors=FALSE)
```

## Summary Statistics

A few values of Loss Adjustment were coded as less than 10. These are replaced with zeros.

```{r}
Vars <- c("IndemnityPaid", "LossAdjustment", "NonEconomicLoss", "InjurySeverity","InsurerType",
          "CountyInjuryOccurred","FinalDisposition","CourtDecision","DateInjuryOccurred","DateInjuryReported")
Data <- SampleData[Vars]
#sum(1*(Data$LossAdjustment>0)*(Data$LossAdjustment<10))
Data$LossAdjustment <- Data$LossAdjustment*(Data$LossAdjustment>10)
#sum(1*(Data$NonEconomicLoss>0)*(Data$NonEconomicLoss<10))
scalefactor <- 1000
Data$IndemnityPaid   <- Data$IndemnityPaid/scalefactor
Data$LossAdjustment  <- Data$LossAdjustment/scalefactor
Data$NonEconomicLoss <- Data$NonEconomicLoss/scalefactor
cnames <- c("Y1", "Y2", "Y3", "x1","x2","x3","x4","x5","Date1","Date2")
Data1 <- Data
colnames(Data1) <- cnames
Data1$PosY2 <- 1*(Data1$Y2>0)
Data1$PosY3 <- 1*(Data1$Y3>0)

```

### Dependent Variables

There are `r length(Data1$Y1)` observations in the dataset. Dependent variables are in units of `r scalefactor`. Note that the minimum of *IndemnityPaid* is `r min(Data1$Y1)` USD.

```{r}
summ2 <- cbind(summary(Data[,1]),summary(Data[,2]),summary(Data[,3]))
colnames(summ2) <- c("Indemnity Paid", "Loss Adjustment", "NonEconomic Loss")
knitr::kable(summ2,digits=0, caption="Dependent Variable Summary Measures")
```


There are `r sum(1-Data1$PosY2)` zeros for *Y2 = LossAdjustment* and `r sum(1-Data1$PosY3)` zeros for *Y3 = NonEconomicLoss*. Here is a table of their joint behavior 

```{r}
table(Data1$PosY2,Data1$PosY3)
```

The variable *IndemnityPaid* has long tails. There are `r Bigdata <- subset(Data, IndemnityPaid > 25000);length(Bigdata$IndemnityPaid)` claims in excess of $25 MM. They are

```{r} 
Bigdata$IndemnityPaid[order(Bigdata$IndemnityPaid)] 
```


Note that the largest claims is `r max(Bigdata$IndemnityPaid)/1000` in millions of USD.

The distribution of these three dependent variables are skewed. Here are summaries of the distribution (after removing zeros from Y2 and Y3).

```{r fig.width=8, fig.height=4}
PosY2data <- subset(Data1, Y2>0)
PosY3data <- subset(Data1, Y3>0)
par(mfrow=c(1,3))
hist(log10(Data1$Y1), xlab = "Logarithmic Indemnity Paid", main="", xlim=c(0,6))
hist(log10(PosY2data$Y2), xlab = "Logarithmic Loss Adjustment", main="", xlim=c(0,6))
hist(log10(PosY3data$Y3), xlab = "Logarithmic NonEconomic Loss", main="", xlim=c(0,6))
SumY1 <- summary(Data1$Y1)
SumY2 <- summary(PosY2data$Y2)
SumY3 <- summary(PosY3data$Y3)
SummaryY <- cbind(SumY1,SumY2,SumY3)
colnames(SummaryY) <- c("Indemnity Paid", "Loss Adjustment", "NonEconomic Loss")
knitr::kable(SummaryY,digits=1, caption="Dependent Variable Summary Measures, Omitting Zeros")

```

### Explanatory Variables

These variables do not help on claims prediction but may help explain claim size. Will need to incorporate the two date variables later.

```{r}
table(Data[,4]); table(Data[,5]); table(Data[,7]);table(Data[,8])

```
Claims by County

```{r}
table(Data[,6])
```

## Relationships

The most interesting variable is **x2**. Self-insurers pay loss adjustment expenses and non-economic damages less often than claims insured by licensed carriers.


```{r warning=FALSE}

Tablexj<-function(xj1, xj2, xj3) {
#  xj1=Data1$x1;xj2=PosY2data$x1;xj3=PosY3data$x1)
r1 <- aggregate(Y1   ~ xj1, data=Data1,    length)$Y1 
r2 <- aggregate(Y1   ~ xj1, data=Data1,    mean)$Y1
r3 <- aggregate(Y1   ~ xj1, data=Data1,    median)$Y1
r4 <- aggregate(PosY2~ xj1, data=Data1,    mean)$PosY2
r5 <- aggregate(Y2   ~ xj2, data=PosY2data,median)$Y2
r5star = r4;rcount=1
for (i in 1:length(r1)){
  if (r4[i] ==0)  r5star[i]<-"NA" else r5star[i]=r5[rcount]
  rcount=rcount+1}
r6 <- aggregate(PosY3~ xj1, data=Data1,    mean)$PosY3
r7 <- aggregate(Y3   ~ xj3, data=PosY3data,median)$Y3
r7star = r6;rcount=1
for (i in 1:length(r1)){
  if (r6[i]==0) r7star[i]<-"NA" else r7star[i]<-r7[rcount]
  rcount=rcount+1}
cbind(r1,r2,r3,r4,as.numeric(r5star),r6,as.numeric(r7star)) 
}
t1 <- Tablexj(Data1$x1,PosY2data$x1,PosY3data$x1)
t2 <- Tablexj(Data1$x2,PosY2data$x2,PosY3data$x2)
#t3 <- Tablexj(Data1$x3,PosY2data$x3,PosY3data$x3)
t4 <- Tablexj(Data1$x4,PosY2data$x4,PosY3data$x4)
t5 <- Tablexj(Data1$x5,PosY2data$x5,PosY3data$x5)
rownames(t1) <- c("x1:Emotional","x1:Permanent:Death","x1:Permanent:Grave","x1:Permanent:Major","x1:Permanent:Minor","x1:Permanent:Significant",
                  "x1:Temporary:Major","x1:Temporary:Minor", "x1:Temporary:Slight")
rownames(t2) <- c("x2:Licensed","x2:Self-Insurer")
rownames(t4) <- c("x4:Disposed of by Court","x4:Settled by parties")
rownames(t5) <- c("x5:Directed verdict for plaintiff","x5:Judgment for the defendant","x5:Judgment for the plaintiff after appeal",
                  "x5:Judgment for the plaintiff","x5:Judgment notwithstanding the verdict for plaintiff",
                  "x5:No Court Proceedings","x5:Other")
#TableMat <- rbind(t1,t2,t3,t4,t5)
TableMat <- rbind(t1,t2,t4,t5)
colnames(TableMat) <-  c("Num Obs","Mean Y1","Median Y1","Prop Y2","Median Pos Y2","Prop Y3", "Median Pos Y3")
knitr::kable(TableMat,digits=3, caption="Explanatory Variables")
```



```{r}
CorMat <- cor(Data[,1:3], method="spearman", use="pairwise.complete.obs")
knitr::kable(CorMat,digits=3, caption="Spearman Correlations")
library(copula)
DataComplete <- subset(Data1, Y2>0 & Y3>0)
splom2(log10(DataComplete[,1:3]), cex=0.8, col.mat="black")

#What is the relationship between indemnity paid and non-economic losses???
sum((DataComplete$Y1<DataComplete$Y3))
sum((DataComplete$Y1==DataComplete$Y3))
sum((DataComplete$Y1>DataComplete$Y3))

```
